---
title: "Analysis Report"
author: "Mehmet Sami Boz"
date: "2025-07-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glue)
```

## Data

Data is downloaded from webpage:
https://data.cityofnewyork.us/Housing-Development/Capital-Project-Schedules-and-Budgets/2xh6-psuq
its name is 
Data.gov: Capital Project Schedules and Budgets.


After examination one can say that each row represents a phase in a project. Phase has values of actual start date,planned end date, actual end date,budget,estimate of actual costs,actual spending.

Some projects have same building several schools for schools there are several buildings. And for some cases for same building same school same project the job entered 5 year plan with different name. And for some rare cases. Eventough they are under the same number in the 5 year plan, project phase conducted again.

Due to these causes we can identify phases of ordinary projects with building identifier, school name, project description,phase name for repeated projects we'll use DSF_numbers and actual start time. geographic district,project type variables can be used for filtering when analyzing.


### Data Cleaning

Some of the value columns have labeled with certain words, following table explains the situation

|label | Project_Status_Name| Project_Phase_Actual_Start_Date| Project_Phase_Planned_End_Date| Project_Phase_Actual_End_Date| Project_Budget_Amount|
|:-----|-------------------:|-------------------------------:|------------------------------:|-----------------------------:|---------------------:|
|PNS   |                4886|                            4863|                           4863|                          4863|                    NA|
|FTK   |                  NA|                              23|                            353|                            23|                   353|
|DIIR  |                  NA|                              NA|                            350|                            NA|                  1058|
|DIIT  |                  NA|                              NA|                             53|                            NA|                    53|
|DOEL  |                  NA|                              NA|                             21|                            NA|                    21|
|DOEP  |                  NA|                              NA|                              5|                            NA|                     5|
|DOER  |                  NA|                              NA|                            661|                            NA|                  1860|
|DOES  |                  NA|                              NA|                           1051|                            NA|                  1051|
|EMER  |                  NA|                              NA|                            119|                            NA|                   119|
|IEH   |                  NA|                              NA|                            694|                            NA|                   694|
|TPL   |                  NA|                              NA|                              9|                            NA|                    12|

When examined I figured out that each label except PNS is about project department, we can understand the situation with following table;

|Project_Type           |Project_Phase_Planned_End_Date |Project_Budget_Amount |
|:----------------------|:------------------------------|:---------------------|
|SCA IEH                |IEH                            |IEH                   |
|DOE - RESOA            |DOER                           |DOER                  |
|DOE - Skilled Trades   |DOES                           |DOES                  |
|Fast Track Projects    |FTK                            |FTK                   |
|DIIT - RESOA           |DIIR                           |DIIR                  |
|Trust For Public Land  |TPL                            |TPL                   |
|SCA Emergency Response |EMER                           |EMER                  |
|DIIT - Project Connect |DIIT                           |DIIT                  |
|DOE - Lead Paint       |DOEL                           |DOEL                  |
|DOE Managed PRE-K      |DOEP                           |DOEP                  |

So we will convert these labels to NA because this labels does not exhibit a information on project apart from the project type which is a distinct column.

However, PNS is different. Eventough there is no information on metadata(or other resources) about PNS, by examining data thoroughly we may conclude that PNS label is about planning. My best guess is PNS stands for "Planned - Not Started"

### NAs


#### Project_Phase_Actual_End_Date

Several columns have NA values most crucial one is Project_Phase_Actual_End_Date.
It has 4533 NA. When examined this is equal to amount of Project_Status of In-progress.
So these data actually is not missing but has not occurred yet so we need to hold the data 

#### Project_Phase_Name

When examined the single observation of Na in this column we saw that by looking other columns values same with other observations this observations Phase name could be Construction.
This implementation can be done because every other column has meaningful values except phase name.

#### Project_Phase_Planned_End_Date

After examination deeply one can find that this columns' NA occurs when project type and if Project status is PNS planned end date is PNS. And if it is completed or In-progress it has value NA.This means these projects starting with no planned end date. So in this case being NA has its own meaning so we cant implement anything that changes mentioned structure or drop these values.

#### Final_Estimate_of_Actual_Costs_Through_End_of_Phase_Amount

Except two cases we can explain these NAs with following relationship; Emergency projects with zero actual spending has no estimated cost except one case. And another case has no estimated cost but it is not an emergency project. The latter case might be dropped out but other 16 cases should be handled as "can't estimate".

#### Total_Phase_Actual_Spending_Amount

In these NAs each phase is completed. Also each phase is a "Construction" phase. No budget is determined for any observation but every case ahs an estimated costs. So we fill the actual cost with estimated cost to prevent information lost we will do this because pahses are completed and there is an estimated amount. So no change is expected and estimation when when a event occurred become observation.

#### DSF_Number(s) 

In this case, each NA occurred at specific project: "LEAD PAINT ABATEMENT". When checked also every "LEAD PAINT ABATEMENT" project has NA DSF numbers. Another thinking point is DSF numbers is about info that connects project to 5 year plan so in terms of project management analysis it has no value. So we can easily drop this column so there will be NA problem also.

### Problematic Values/observations

If a phase has time or budget variables as group of columns it is okay for me we don't seek both of them to be exist but if none is available then no need to hold that observation.
Giving aggregated statistics is excluded from above statement. So we can only drop values after creating KPI columns.

In Project Phase names "CM,Art,F&E","CM","F&E","CM,F&E" is merged and named "CM,F&E". Because these are not so different from each other and have small number of observations.


### EDA

```{r,echo=FALSE}
load("../data/data_ready.Rdata")
full_data <- data %>% mutate(Project_Phase_Actual_Start_Date = mdy(Project_Phase_Actual_Start_Date),
                             Project_Phase_Planned_End_Date = mdy(Project_Phase_Planned_End_Date),
                             Project_Phase_Actual_End_Date = mdy(Project_Phase_Actual_End_Date),
                             Project_Budget_Amount = as.numeric(Project_Budget_Amount))
```


#### Categorical variables

```{r,echo=FALSE}
janitor::tabyl(full_data,Project_Phase_Name,Project_Status_Name)
```

We see that majority of Construction and Furniture & Equipment  phases are either in progress or under planning. On th eother hand, Design and Scope phases are mostly done.

#### Numeric and Date variables

in each cell we are seeing mean - median (std.dev) min- max

<table>
<tbody>
  <tr>
   <td style="text-align:left;"> Project_Phase_Actual_Start_Date </td>
   <td style="text-align:left;"> 2022-10-08 - 2023-06-14 (871.6)
2003-09-12 - 2025-03-31 </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Project_Phase_Planned_End_Date </td>
   <td style="text-align:left;"> 2023-05-17 - 2023-12-18 (949)
2003-09-12 - 2029-02-01 </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Project_Phase_Actual_End_Date </td>
   <td style="text-align:left;"> 2023-03-06 - 2023-10-18 (838.6)
2003-09-12 - 2025-03-31 </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Project_Budget_Amount </td>
   <td style="text-align:left;"> 2026553.9 - 125330 (8197260.6)
0 - 202994750 </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Final_Estimate_of_Actual_Costs_Through_End_of_Phase_Amount </td>
   <td style="text-align:left;"> 1304951.3 - 125000 (6322911.5)
0 - 199619196 </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Total_Phase_Actual_Spending_Amount </td>
   <td style="text-align:left;"> 463290.7 - 8175 (3071916.9)
0 - 195252948 </td>
  </tr>
</tbody>
</table>



From the data we are seeing that most of the projects is gathered around 2022-2023.
In terms of spending columns mean and median difference shows us the large projects effect on the dataset as outliers.

## KPI

I created delay_ratio and cost_override as my key performance indicators.

### Delay ratio 

$$
\text{Completion_time} = \text{Project_Phase_Actual_End_Date}-\text{Project_Phase_Actual_Start_Date} \\
\text{Planned_time} = \text{Project_Phase_Planned_End_Date}-\text{Project_Phase_Actual_Start_Date} \\
\text{Delay_ratio} = \frac{\text{Completion_time}}{\text{Planned_time}} 
$$
In this ratio, phases with one done in exactly planned time, ratios lower than one represent quick completions. Finally more than one tells us that project phases with delays. Higher the value more the delay in term "times": ratio 2 means phase completed 2 times the amount planned.

Two observations have negative delay ratios means that actual start date is after planned end date which is a mistake most probably project phase is postponed due to previous phases lags. Because we cannot analyze it we drop these two values.
Another 16 observation is removed temporarily to be able model Delay ratio. These observation have zero delay ratio which means they finished the phase same day they started.

Another remark to keep in mind is because this is created on ended project phase this column don't give any insight about planning (PNS) or in progress phases.
Also because scope and design is early steps of a project each completed construction phase have completed scope and design inverse is obliviously not true. So analysis contains these values much more than others.

### Cost Ratio

$$
\text{Cost_override}= \frac{\text{Total_Phase_Actual_Spending_Amount} - \text{Project_Budget_Amount}}{\text{Project_Budget_Amount}}
$$

If actual spending is more than the budget cost ratio will be positive. If the value is negative then project phase is in the safe zone.

There are some (namely 437) zero budget phases with this formula that zeros becomes Infinity. So we drop these cases to analyze further.

Some project phases multiplied budgets so we see so big values in this ratio like 141. there are 10 observation with ratio above 10 (actual spending is ten times budget amount).

## Modelling

### Statistical Models

#### Duration

To model and understand delays in this dataset we modeled deep information about analysis can be found in the app.

#### Cost

Same as above for Costs paths I built a statistical model to understand relationship between other variables and costs. Deep explanation including interpretation can be found in the shiny app.

## Shiny

To present findings to management and gather insight quickly, a Shiny app is designed.

### Panels

#### Duration Analysis

In this panel, one can found my analysis about delays

#### Cost Analysis

In this panel, cost model especially with respect to budget of the project phase. Both duration and cost analysis made with unique observation of phases. Each project consists of phases as mentioned in the data paragraph atomic core of this dataset is project phases not projects.

#### Observation

In this panels I built two warning and observe dashboard. First one is by project and second one is by phase. Former one is aggregated from phases and represent filters coming thanks to this aggregation. Then with a simple reactive use we can get total warnings also users can examine the raw data and examine the warnings directly. In project level there are three warnings. These are;  Delay and Cost Overrun where both cost and delay exceeded threshold. Cost Overrun: triggered when actual spending is more than project budget. Delay: where actual end date is later than planned end date of the project.

Second dashboard is again for observation and made in level phases. One can examine the warnings with respect to phase names and phase status.Users also filter projects according to their start date. In this panel there are 5 warnings details can be found in the dictionary. These warnings are as follows: Early Cost overrun warning, Realized Cost Overrun Warning, Realized Delay Warning, Early Delay Warning, Critical Cost Overrun and Delay Warning

## AI


*Used for recommendation:*

- “I need to make a terms dictionary for my project in GitHub I'm using R and It will be in rmd file any recommendations?”

- “how to design a warning mechanism for project manager?”
  -“what if for a sample project I have static data”

- “I want a checkbox says show data if so data tabşe appears but if not then nothing”

- “in shiny I will put two input as slider but each one will be activated if related switch is turned on how to do that”

- “how to pie chart in ggplot”

*Error handling:*

- “is theme element is not an exported object from namespace ggplot2 when thematic shiny why I am seeing this error”

*To learn something:*

- “how to make css files for shiny can you give me a deep introduction”


And as discussed in the interview I never copy paste anything from other to AI as my future working procedure will be like that I asked questions learned something and applied if I needed a helper. Also I don’t write down my data as well because in future my data will be classified.

Other than AI, my main sources are Shiny webpage, and Hadley Wickam’s Mastering Shiny books. Couldn’t have time for the book Outstanding User Interfaces with Shiny (for CSS) but it seems a nice source.
For some errors I reached stackoverflow but couldn’t anything useful. That’s why my plots do not react to dark mode in the shiny app.

## Dictionary

**PNS**
 Term found in dataset guessed as it stands for planning 

**Early Cost Overrun Warning**
 this warning determined by comparing estimate spending is higher than budget. Phase level. 

**Realized Cost Overrun Warning**
 this warning determined by comparing actual spending is higher than the budget. Phase level. 

**Realized Delay Warning**
 Actual end date is later than planned end date. Phase level 

**Early Delay Warning**
 For In-progress phases that planned end date is later than dataset time which is last end date observed: end of March 2025. Phase level 

**Critical Cost Overrun and Delay Warning**
 Raised if both Cost and delay warnings are active.Phase level. 

**Project Geographic District**
 District where building is located 

**Project Building Identifier**
 Building ID 

**Project School Name**
 School name 

**Project Type**
 Type of project based on funding 

**Project Description**
 Component(s) of work to be done 

**Project Phase Name**
 Examples Scope , Design & Construction 

**Project Status Name**
 Within the phase more detail, i.e. In-Progress, Hold 

**Project Phase Actual Start Date**
 Date phase actually started 

**Project Phase Planned End Date**
 Date phase originally scheduled to be completed 

**Project Phase Actual End Date**
 Date phase actually completed 

**Project Budget Amount**
 Award base budget by phase 

**Final Estimate of Actual Costs Through End of Phase Amount**
 Current projected final estimate at completion of project by phase 

**Total Phase Actual Spending Amount**
 Actual cumulative expenditures by phase 

**DSF Number(s)**
 Number used to identify project in Five Year Plan 

**Phase**
 Each seperate part of a project 

**KPI**
 Key performance indicator 

**EDA**
 Explotary data analysis 

**CM,F&E**
 Construction Management, Furniture & Equipment – (Guessed) 

**Construction**
 Main construction phase – (Guessed) 

**Design**
 Architectural and engineering design – (Guessed) 

**Purch & Install**
 Purchase and Installation – (Guessed) 

**Scope**
 Scoping and pre-design – (Guessed) 